```{r}
library(dplyr)
library(lubridate)
library(forecast)
library(tseries)
library(ggplot2)
```
```{r}
df_fruitveg <- readr::read_csv("data/cleaned_fruit_veg.csv", show_col_types = FALSE) %>%
  mutate(
    date = as.Date(date),
    price = as.numeric(price)
  ) %>%
  dplyr::select(date, price) %>%
  arrange(date)

weekly_index <- df_fruitveg %>%
  dplyr::select(date) %>%
  arrange(date)
```
```{r}
df_price_weekly <- df_fruitveg %>%
  dplyr::select(date, price) %>%
  dplyr::arrange(date)
```
```{r}
arima_series <- df_price_weekly %>%
  dplyr::group_by(date) %>%
  dplyr::summarise(price = mean(price, na.rm = TRUE), .groups = "drop") %>%
  dplyr::arrange(date) %>%
  dplyr::filter(!is.na(price))

price_ts <- ts(
  arima_series$price,
  start     = c(lubridate::year(min(arima_series$date)),
                lubridate::week(min(arima_series$date))),
  frequency = 52
)

cat("Series length:", length(price_ts), "weeks\n")
cat("Date range:", as.character(range(arima_series$date)), "\n")
```
```{r}
cat("\n── ADF Test: Levels ──\n")
adf_level <- tseries::adf.test(price_ts)
print(adf_level)

cat("\n── ADF Test: First Difference ──\n")
adf_diff <- tseries::adf.test(diff(price_ts))
print(adf_diff)

# Determine differencing order
d <- if (adf_level$p.value < 0.05) 0 else 1
cat("\nDifferencing order selected: d =", d, "\n")
```
```{r}
price_diff <- if (d == 1) diff(price_ts) else price_ts

par(mfrow = c(2, 1), mar = c(4, 4, 2, 1))
acf(price_diff,  lag.max = 52, main = "ACF — Differenced Weekly Price")
pacf(price_diff, lag.max = 52, main = "PACF — Differenced Weekly Price")
par(mfrow = c(1, 1))
```
```{r}
cat("\n── Auto ARIMA (stepwise) ──\n")
fit_auto <- forecast::auto.arima(
  price_ts,
  d           = d,
  stepwise    = TRUE,
  approximation = FALSE,
  trace       = TRUE
)
cat("\nSelected model:\n")
print(fit_auto)
cat("\nAIC:", AIC(fit_auto), "| BIC:", BIC(fit_auto), "\n")
```
```{r}
cat("\n── SARIMA(1,1,1)(1,1,1)[52] ──\n")
fit_sarima <- tryCatch(
  forecast::Arima(price_ts,
                  order    = c(1, d, 1),
                  seasonal = list(order = c(1, 1, 1), period = 52)),
  error = function(e) { cat("SARIMA failed:", e$message, "\n"); NULL }
)
if (!is.null(fit_sarima)) {
  print(fit_sarima)
  cat("AIC:", AIC(fit_sarima), "| BIC:", BIC(fit_sarima), "\n")
}
```
```{r}
cat("\n── STL + ARIMA (recommended for weekly data) ──\n")
fit_stl <- forecast::stlf(
  price_ts,
  method  = "arima",
  h       = 52,
  robust  = TRUE
)
cat("STL decomposition + ARIMA inner model:\n")
print(fit_stl$model)
cat("AIC:", AIC(fit_stl$model), "\n")
```
```{r}
cat("\n── Accuracy Comparison ──\n")
acc_auto   <- forecast::accuracy(fit_auto)
acc_stl    <- forecast::accuracy(fit_stl)

cat("\nAuto ARIMA:\n");  print(round(acc_auto, 4))
cat("\nSTL + ARIMA:\n"); print(round(acc_stl, 4))

if (!is.null(fit_sarima)) {
  acc_sarima <- forecast::accuracy(fit_sarima)
  cat("\nSARIMA:\n"); print(round(acc_sarima, 4))
}

model_comparison <- data.frame(
  Model = c("Auto ARIMA", "STL + ARIMA",
            if (!is.null(fit_sarima)) "SARIMA(1,1,1)(1,1,1)[52]" else NULL),
  RMSE  = c(acc_auto[1,"RMSE"],  acc_stl[1,"RMSE"],
            if (!is.null(fit_sarima)) acc_sarima[1,"RMSE"] else NULL),
  MAE   = c(acc_auto[1,"MAE"],   acc_stl[1,"MAE"],
            if (!is.null(fit_sarima)) acc_sarima[1,"MAE"] else NULL),
  MASE  = c(acc_auto[1,"MASE"],  acc_stl[1,"MASE"],
            if (!is.null(fit_sarima)) acc_sarima[1,"MASE"] else NULL),
  AIC   = c(AIC(fit_auto), AIC(fit_stl$model),
            if (!is.null(fit_sarima)) AIC(fit_sarima) else NULL)
)
cat("\n── Model Comparison Table ──\n")
print(round(model_comparison[, -1], 4) |> cbind(Model = model_comparison$Model, ...=_))
```
```{r}
h <- 52

fc_auto <- forecast::forecast(fit_auto, h = h)
fc_stl  <- fit_stl  # already forecasted in stlf()

# Best model forecast plot (ggplot)
fc_df <- data.frame(
  date = arima_series$date[length(arima_series$date)] + (1:h) * 7,
  mean = as.numeric(fc_auto$mean),
  lo80 = as.numeric(fc_auto$lower[, 1]),
  hi80 = as.numeric(fc_auto$upper[, 1]),
  lo95 = as.numeric(fc_auto$lower[, 2]),
  hi95 = as.numeric(fc_auto$upper[, 2])
)

p_arima <- ggplot() +
  geom_line(data = arima_series, aes(x = date, y = price),
            colour = "steelblue", linewidth = 0.6) +
  geom_ribbon(data = fc_df, aes(x = date, ymin = lo95, ymax = hi95),
              fill = "darkorange", alpha = 0.15) +
  geom_ribbon(data = fc_df, aes(x = date, ymin = lo80, ymax = hi80),
              fill = "darkorange", alpha = 0.25) +
  geom_line(data = fc_df, aes(x = date, y = mean),
            colour = "darkorange", linewidth = 0.8, linetype = "dashed") +
  labs(
    title    = paste("ARIMA Forecast —", fit_auto),
    subtitle = "52-week ahead forecast with 80% and 95% prediction intervals",
    x = NULL, y = "Price Index",
    caption  = "Auto ARIMA selected model. Shaded bands: 80% (dark) and 95% (light) intervals."
  ) +
  theme_minimal(base_size = 12) +
  theme(plot.title = element_text(face = "bold"))

print(p_arima)

# STL forecast plot
fc_stl_df <- data.frame(
  date = arima_series$date[length(arima_series$date)] + (1:h) * 7,
  mean = as.numeric(fc_stl$mean),
  lo80 = as.numeric(fc_stl$lower[, 1]),
  hi80 = as.numeric(fc_stl$upper[, 1]),
  lo95 = as.numeric(fc_stl$lower[, 2]),
  hi95 = as.numeric(fc_stl$upper[, 2])
)

p_stl <- ggplot() +
  geom_line(data = arima_series, aes(x = date, y = price),
            colour = "steelblue", linewidth = 0.6) +
  geom_ribbon(data = fc_stl_df, aes(x = date, ymin = lo95, ymax = hi95),
              fill = "darkgreen", alpha = 0.15) +
  geom_ribbon(data = fc_stl_df, aes(x = date, ymin = lo80, ymax = hi80),
              fill = "darkgreen", alpha = 0.25) +
  geom_line(data = fc_stl_df, aes(x = date, y = mean),
            colour = "darkgreen", linewidth = 0.8, linetype = "dashed") +
  labs(
    title    = "STL + ARIMA Forecast — UK Weekly Fruit & Veg Prices",
    subtitle = "52-week ahead forecast with 80% and 95% prediction intervals",
    x = NULL, y = "Price Index",
    caption  = "STL decomposition handles weekly seasonality; ARIMA fits residuals."
  ) +
  theme_minimal(base_size = 12) +
  theme(plot.title = element_text(face = "bold"))

print(p_stl)
```
```{r}
# ── STEP 8: Residual diagnostics ─────────────────────────────────────────────
cat("\n── Residual Diagnostics: Auto ARIMA ──\n")
forecast::checkresiduals(fit_auto)

cat("\n── Residual Diagnostics: STL + ARIMA ──\n")
forecast::checkresiduals(fit_stl)

# Ljung-Box explicitly
lb_auto <- Box.test(residuals(fit_auto), lag = 20, type = "Ljung-Box")
lb_stl  <- Box.test(residuals(fit_stl),  lag = 20, type = "Ljung-Box")

cat("\nLjung-Box p-value — Auto ARIMA:", round(lb_auto$p.value, 4),
    ifelse(lb_auto$p.value > 0.05, "✓ White noise", "✗ Autocorrelation remains"), "\n")
cat("Ljung-Box p-value — STL+ARIMA: ", round(lb_stl$p.value, 4),
    ifelse(lb_stl$p.value > 0.05, "✓ White noise", "✗ Autocorrelation remains"), "\n")
```
**ARIMA Forecasting — UK Weekly Fruit & Vegetable Prices**

---

**Data and Stationarity**

Following aggregation to a single mean price per week, the series comprises 527 observations spanning January 2015 to February 2026 — a clean weekly aggregate, resolving the 8,308-observation issue present in the Holt-Winters run. The ADF test on levels returned a Dickey-Fuller statistic of −5.92 (p < 0.01), rejecting the null of a unit root and indicating the series is stationary in levels. This is somewhat unusual for a price series but plausible here given the relatively short 11-year window and the mean-reverting nature of agricultural commodity prices around a long-run trend. Accordingly, no differencing was applied (d = 0), which `auto.arima` confirmed in its model search.

---

**Model Selection**

The `auto.arima` search evaluated 22 candidate specifications across ARIMA and seasonal ARIMA combinations, using AICc as the selection criterion. The selected model was **ARIMA(2,0,0) with non-zero mean** — a second-order autoregressive process with no moving average component and no differencing. The estimated coefficients are AR(1) = 0.677 (s.e. 0.042) and AR(2) = 0.230 (s.e. 0.042), both significant, with a long-run mean of 1.329. The positive and decaying AR coefficients indicate that this week's price is strongly influenced by the previous two weeks, with the effect attenuating over time — a classic persistence structure for commodity prices. The model's AIC of −18.75 was the lowest of all candidates evaluated.

Three models were ultimately estimated and compared:

| Model | RMSE | MAE | MASE | AIC |
|---|---|---|---|---|
| Auto ARIMA(2,0,0) | 0.2356 | 0.1295 | 0.4888 | −18.75 |
| SARIMA(1,0,1)(1,1,1)[52] | 0.2355 | 0.1346 | 0.5078 | 110.38 |
| STL + ARIMA(4,1,2) | 0.2898 | 0.1681 | 0.6342 | 206.33 |

The Auto ARIMA model achieves the best AIC by a substantial margin and the best MAE and MASE. The SARIMA model matches it almost exactly on RMSE (0.2355 vs 0.2356) but carries a much higher AIC due to the additional seasonal parameters — suggesting the seasonal structure it captures does not justify the added complexity. The STL + ARIMA is the weakest on all in-sample accuracy metrics, though this is expected since STL first removes the trend-seasonal component before ARIMA fits the residuals, which limits in-sample fit.

All three MASE values are well below 1.0, confirming that every model substantially outperforms a naïve random walk forecast — a meaningful benchmark for evaluation.

---

**Residual Diagnostics**

The Ljung-Box test results require careful interpretation because two different tests were applied. The `checkresiduals()` function uses a lag of approximately 2× the model order plus the seasonal period (104 lags total), which is a stringent test. Under this specification, the Auto ARIMA residuals show significant autocorrelation (Q* = 139.37, df = 102, p = 0.008) and STL + ARIMA residuals show highly significant autocorrelation (Q* = 248.44, df = 98, p ≈ 0). This indicates residual structure remains at longer lags — almost certainly the seasonal pattern at lag 52, which neither model fully captures.

However, when the Ljung-Box test is applied at the more conventional lag of 20, both models pass cleanly: Auto ARIMA (p = 0.619) and STL + ARIMA (p = 0.442) both indicate white noise residuals at short to medium lags. This distinction matters for interpretation — the models capture the short-run autocorrelation structure well, but leave some long-run seasonal signal in the residuals. For a 52-week forecast horizon this is a known limitation of ARIMA applied to weekly data, and is the reason STL decomposition is generally recommended for weekly series despite its weaker in-sample fit.

---

**Forecast**

The ARIMA(2,0,0) 52-week ahead forecast projects a gradual mean reversion toward the long-run level of approximately 1.33, with prediction intervals widening smoothly over the horizon. Because d = 0 and the AR coefficients sum to less than 1 (0.677 + 0.230 = 0.907), the forecast is stationary and will always revert to the mean rather than trending indefinitely — appropriate behaviour for an agricultural price index. The 95% prediction interval spans roughly ±0.5 price index points at the 52-week horizon.

The STL + ARIMA forecast captures the seasonal pattern explicitly in the forecast trajectory, producing a more textured forward path that rises and falls with the estimated 52-week seasonal cycle. This is arguably more realistic for a food price series with genuine seasonal agricultural patterns, even though its in-sample fit is weaker.

---

**Model Recommendation**

For point forecast accuracy and parsimony, **ARIMA(2,0,0)** is the best performing model by AIC, MAE, and MASE, and its residuals are well-behaved at the standard lag-20 threshold. However, for a report communicating forecast uncertainty over a full year ahead, the **STL + ARIMA** forecast is more interpretable and economically defensible — it produces a seasonally-shaped forecast path rather than a flat mean reversion, and the STL decomposition framework explicitly handles the weekly frequency that causes ETS and standard ARIMA to ignore seasonality entirely. The two models are best presented together: ARIMA(2,0,0) as the statistically optimal specification, and STL + ARIMA as the seasonally-aware forecast for stakeholder communication.