---
title: "Granger Causality Analysis"
author: "Harleen Buttar, Danilo, Minhao Zhang"
output: 
  html_document:
    toc: true
    toc_float: true
    theme: cosmo
---

```{r}
library(readr)
library(readxl)
library(dplyr)
library(lubridate)
library(zoo)
library(lmtest)
library(vars)
library(tidyr)
```

```{r}
# Find a column name by regex pattern (case-insensitive)
pick_col <- function(df, pattern, required = TRUE){
  nms <- names(df)
  hit <- nms[grepl(pattern, nms, ignore.case = TRUE)]
  if(length(hit) == 0){
    if(required) stop(paste("Could not find a column matching pattern:", pattern))
    return(NA_character_)
  }
  hit[1]
}

# Convert low-frequency series (monthly/quarterly) to weekly using LOCF on weekly date index
to_weekly_locf <- function(weekly_index, lowfreq_df, value_col) {

  lowfreq_df_clean <- lowfreq_df %>%
    dplyr::select(date, !!rlang::sym(value_col)) %>%
    dplyr::group_by(date) %>%                 # ← CRITICAL
    dplyr::summarise(
      value = mean(.data[[value_col]], na.rm = TRUE),
      .groups = "drop"
    )

  weekly_index %>%
    dplyr::left_join(lowfreq_df_clean, by = "date") %>%
    dplyr::arrange(date) %>%
    dplyr::mutate(
      value = zoo::na.locf(value, na.rm = FALSE)
    ) %>%
    dplyr::rename(!!value_col := value)
}

# Aggregate any weekly data frame to month/quarter by mean
to_monthly_mean <- function(df, value_col) {
  df %>%
    dplyr::mutate(date = lubridate::floor_date(date, "month")) %>%
    dplyr::group_by(date) %>%
    dplyr::summarise(
      !!value_col := mean(.data[[value_col]], na.rm = TRUE),
      .groups = "drop"
    ) %>%
    dplyr::arrange(date)
}

to_quarterly_mean <- function(df, value_cols){
  df %>%
    dplyr::mutate(date = lubridate::floor_date(date, "quarter")) %>%
    dplyr::group_by(date) %>%
    dplyr::summarise(across(all_of(value_cols), ~mean(.x, na.rm = TRUE)), .groups = "drop") %>%
    dplyr::arrange(date)
}

# Pairwise Granger (x -> y) on first differences
granger_pairwise <- function(df, y, x, lags){
  d <- df %>%
    arrange(date) %>%
    mutate(
      dy = .data[[y]] - lag(.data[[y]]),
      dx = .data[[x]] - lag(.data[[x]])
    ) %>%
    drop_na(dy, dx)
  
  gt <- lmtest::grangertest(dy ~ dx, order = lags, data = d)
  p <- gt$`Pr(>F)`[2]
  tibble(predictor = x, lags = lags, p_value = p)
}

# Joint Granger using VAR and vars::causality:
# tests whether ALL predictors jointly Granger-cause y.
granger_joint <- function(df, y, xs, lags){

  # First differences for stationarity
  d <- df %>%
    dplyr::arrange(date) %>%
    dplyr::mutate(
      dplyr::across(
        dplyr::all_of(c(y, xs)),
        ~ .x - dplyr::lag(.x),
        .names = "d_{.col}"
      )
    ) %>%
    tidyr::drop_na()

  dy  <- paste0("d_", y)
  dxs <- paste0("d_", xs)

  # VAR needs at least y + one predictor
  var_df <- d %>%
    dplyr::select(dplyr::all_of(c(dy, dxs)))

  if (ncol(var_df) < 2) {
    stop("Need at least y and one predictor for joint test.")
  }

  fit <- vars::VAR(var_df, p = lags, type = "const")

  caus <- vars::causality(fit, cause = dxs)

  tibble::tibble(
    test    = "Joint (all predictors)",
    lags    = lags,
    p_value = caus$Granger$p.value
  )
}
```
```{r}
df_fruitveg <- readr::read_csv("data/cleaned_fruit_veg.csv", show_col_types = FALSE) %>%
  mutate(
    date = as.Date(date),
    price = as.numeric(price)
  ) %>%
  dplyr::select(date, price) %>%
  arrange(date)

weekly_index <- df_fruitveg %>%
  dplyr::select(date) %>%
  arrange(date)
```
```{r}
# Create weekly price dataframe (df_fruitveg is already weekly)
df_price_weekly <- df_fruitveg %>%
  dplyr::select(date, price) %>%
  dplyr::arrange(date)
```
```{r}
df_fuel_weekly <- bind_rows(
  read_csv("data/weekly_road_fuel_prices_2003_to_2017.csv", show_col_types = FALSE),
  read_csv("data/weekly_road_fuel_prices_2018_to_now.csv", show_col_types = FALSE)
)

# Date column is "Date" in your files; parse UK format.
fuel_date_col <- pick_col(df_fuel_weekly, "^Date$")
ulsp_col <- pick_col(df_fuel_weekly, "ULSP")

df_fuel_weekly <- df_fuel_weekly %>%
  transmute(
    date = dmy(.data[[fuel_date_col]]),
    fuel_price = as.numeric(.data[[ulsp_col]])
  ) %>%
  arrange(date)
```
```{r}
library(readxl)
library(dplyr)
library(tidyr)
library(lubridate)

raw_cpi <- read_excel(
  "data/consumerpriceinflationdetailedreferencetables.xlsx",
  sheet = "Table 15a, 15b, 15c",
  skip = 5
)

df_cpi_monthly <- raw_cpi %>%
  dplyr::filter(grepl("^[0-9]{4}$", as.character(`...2`))) %>%
  
  dplyr::select(
    year = `...2`,
    Jan, Feb, Mar, Apr, May, Jun, Jul, Aug, Sep, Oct, Nov, Dec
  ) %>%
  
  dplyr::mutate(year = as.integer(year)) %>%
  
  tidyr::pivot_longer(
    cols = Jan:Dec,
    names_to = "month",
    values_to = "cpi"
  ) %>%
  
  # Clean ONS placeholders explicitly
  dplyr::mutate(
    cpi = dplyr::na_if(cpi, ".."),
    cpi = dplyr::na_if(cpi, "-"),
    cpi = as.numeric(cpi)
  ) %>%
  
  dplyr::filter(!is.na(cpi)) %>%
  
  dplyr::mutate(
    date = as.Date(paste(year, month, "01"), format = "%Y %b %d")
  ) %>%
  
  dplyr::select(date, cpi) %>%
  dplyr::arrange(date)
```
```{r}
raw_api <- read_csv(
  "data/API_20260129.csv",
  show_col_types = FALSE
)
```

```{r}
df_api_monthly <- raw_api %>%
  dplyr::transmute(
    date = as.Date(.data[["date"]]),   # ISO date → Date
    api  = as.numeric(.data[["index"]])
  ) %>%
  dplyr::filter(!is.na(date), !is.na(api)) %>%
  dplyr::arrange(date)
```
```{r}
excel_sheets("data/GBFertiliserPriceSeries.xlsx")
```
```{r}
library(readxl)
library(dplyr)
library(readr)

raw_fert <- read_excel(
  "data/GBFertiliserPriceSeries.xlsx",
  sheet = "GB Fertiliser Price Series ",
  skip = 13,
  col_types = "text"
)

df_fert_monthly <- raw_fert %>%
  dplyr::transmute(
    # Excel serial date → Date
    date = as.Date(as.numeric(Month), origin = "1899-12-30"),

    # Choose ONE fertiliser series (UK-produced AN)
    fertiliser = parse_number(`AN – UK produced (34.5% N)`)
  ) %>%
  dplyr::filter(!is.na(date), !is.na(fertiliser)) %>%
  dplyr::arrange(date)

# sanity checks
summary(df_fert_monthly)
range(df_fert_monthly$date)
nrow(df_fert_monthly)
```

```{r}
library(dplyr)
library(readr)
library(zoo)

raw_sppi <- read_csv(
  "data/series-210226.csv",
  col_names = c("period", "value"),
  show_col_types = FALSE
)

df_sppi_quarterly <- raw_sppi %>%
  dplyr::filter(grepl("^[0-9]{4} Q[1-4]$", period)) %>%
  
  dplyr::mutate(
    sppi = as.numeric(value),
    date = as.Date(zoo::as.yearqtr(period, format = "%Y Q%q"))
  ) %>%
  
  dplyr::select(date, sppi) %>%
  dplyr::filter(!is.na(sppi)) %>%
  dplyr::arrange(date)
```


```{r}
df_cpi_monthly <- df_cpi_monthly %>%
  dplyr::group_by(date) %>%
  dplyr::summarise(cpi = mean(cpi, na.rm = TRUE), .groups = "drop")

df_api_monthly <- df_api_monthly %>%
  dplyr::group_by(date) %>%
  dplyr::summarise(api = mean(api, na.rm = TRUE), .groups = "drop")

df_fert_monthly <- df_fert_monthly %>%
  dplyr::group_by(date) %>%
  dplyr::summarise(fertiliser = mean(fertiliser, na.rm = TRUE), .groups = "drop")
```
```{r}
# Monthly target and weekly fuel aggregated
df_price_monthly <- to_monthly_mean(df_fruitveg, "price")
df_fuel_monthly  <- to_monthly_mean(df_fuel_weekly, "fuel_price")

# Convert quarterly SPPI to monthly via LOCF on monthly grid
monthly_index <- df_price_monthly %>%
  dplyr::select(date) %>%
  dplyr::arrange(date)
df_sppi_monthly <- monthly_index %>%
  dplyr::left_join(df_sppi_quarterly, by = "date") %>%
  dplyr::arrange(date) %>%
  dplyr::mutate(sppi = zoo::na.locf(sppi, na.rm = FALSE))

df_monthly <- df_price_monthly %>%
  dplyr::left_join(df_fuel_monthly, by = "date") %>%
  dplyr::left_join(df_cpi_monthly, by = "date") %>%
  dplyr::left_join(df_api_monthly, by = "date") %>%
  dplyr::left_join(df_fert_monthly, by = "date") %>%
  dplyr::left_join(
    df_sppi_monthly %>% dplyr::select(date, sppi),
    by = "date"
  ) %>%
  dplyr::arrange(date)
```
```{r}
df_price_quarterly <- to_quarterly_mean(df_fruitveg, c("price"))
df_fuel_quarterly  <- to_quarterly_mean(df_fuel_weekly, c("fuel_price"))

# Convert monthly CPI/API/Fert to quarterly by mean
df_cpi_quarterly  <- to_quarterly_mean(df_cpi_monthly, c("cpi"))
df_api_quarterly  <- to_quarterly_mean(df_api_monthly, c("api"))
df_fert_quarterly <- to_quarterly_mean(df_fert_monthly, c("fertiliser"))

df_quarterly <- df_price_quarterly %>%
  left_join(df_fuel_quarterly,  by = "date") %>%
  left_join(df_cpi_quarterly,   by = "date") %>%
  left_join(df_api_quarterly,   by = "date") %>%
  left_join(df_fert_quarterly,  by = "date") %>%
  left_join(df_sppi_quarterly,  by = "date") %>%
  arrange(date)
```
```{r}
# ── Fuel alignment (Monday → Friday via week-snapping) ───────────────────────
df_fuel_weekly <- df_fuel_weekly %>%
  dplyr::filter(date >= min(df_price_weekly$date))

df_fuel_weekly_aligned <- df_fuel_weekly %>%
  dplyr::mutate(week = lubridate::floor_date(date, "week", week_start = 5))

weekly_index_snapped <- weekly_index %>%
  dplyr::mutate(week = lubridate::floor_date(date, "week", week_start = 5))

df_fuel_weekly_aligned <- weekly_index_snapped %>%
  dplyr::left_join(
    df_fuel_weekly_aligned %>% dplyr::select(week, fuel_price),
    by = "week"
  ) %>%
  dplyr::select(date, fuel_price) %>%
  dplyr::arrange(date)

# ── Deduplicate source dataframes ────────────────────────────────────────────
df_sppi_quarterly <- df_sppi_quarterly %>%
  dplyr::group_by(date) %>%
  dplyr::summarise(sppi = mean(sppi, na.rm = TRUE), .groups = "drop") %>%
  dplyr::arrange(date)

df_cpi_monthly <- df_cpi_monthly %>%
  dplyr::group_by(date) %>%
  dplyr::summarise(cpi = mean(cpi, na.rm = TRUE), .groups = "drop") %>%
  dplyr::arrange(date)

df_api_monthly <- df_api_monthly %>%
  dplyr::group_by(date) %>%
  dplyr::summarise(api = mean(api, na.rm = TRUE), .groups = "drop") %>%
  dplyr::arrange(date)

df_fert_monthly <- df_fert_monthly %>%
  dplyr::group_by(date) %>%
  dplyr::summarise(fertiliser = mean(fertiliser, na.rm = TRUE), .groups = "drop") %>%
  dplyr::arrange(date)

# ── LOCF fill to weekly ───────────────────────────────────────────────────────
weekly_index <- df_price_weekly %>%
  dplyr::select(date) %>%
  dplyr::arrange(date)

df_cpi_weekly  <- to_weekly_locf(weekly_index, df_cpi_monthly,    "cpi")
df_api_weekly  <- to_weekly_locf(weekly_index, df_api_monthly,    "api")
df_fert_weekly <- to_weekly_locf(weekly_index, df_fert_monthly,   "fertiliser")
df_sppi_weekly <- to_weekly_locf(weekly_index, df_sppi_quarterly, "sppi")

# ── Deduplicate outputs ───────────────────────────────────────────────────────
df_cpi_weekly  <- df_cpi_weekly %>%
  dplyr::group_by(date) %>%
  dplyr::summarise(cpi = mean(cpi, na.rm = TRUE), .groups = "drop")

df_api_weekly  <- df_api_weekly %>%
  dplyr::group_by(date) %>%
  dplyr::summarise(api = mean(api, na.rm = TRUE), .groups = "drop")

df_fert_weekly <- df_fert_weekly %>%
  dplyr::group_by(date) %>%
  dplyr::summarise(fertiliser = mean(fertiliser, na.rm = TRUE), .groups = "drop")

df_sppi_weekly <- df_sppi_weekly %>%
  dplyr::group_by(date) %>%
  dplyr::summarise(sppi = mean(sppi, na.rm = TRUE), .groups = "drop")

# Check every input immediately before joining
cat("price:", nrow(df_price_weekly), "dupes:", sum(duplicated(df_price_weekly$date)), "\n")
cat("fuel_aligned:", nrow(df_fuel_weekly_aligned), "dupes:", sum(duplicated(df_fuel_weekly_aligned$date)), "\n")
cat("cpi:", nrow(df_cpi_weekly), "dupes:", sum(duplicated(df_cpi_weekly$date)), "\n")
cat("api:", nrow(df_api_weekly), "dupes:", sum(duplicated(df_api_weekly$date)), "\n")
cat("fert:", nrow(df_fert_weekly), "dupes:", sum(duplicated(df_fert_weekly$date)), "\n")
cat("sppi:", nrow(df_sppi_weekly), "dupes:", sum(duplicated(df_sppi_weekly$date)), "\n")

# Join one at a time and check after each step
df_weekly <- df_price_weekly
cat("after price:", nrow(df_weekly), "\n")

df_weekly <- df_weekly %>% dplyr::left_join(df_fuel_weekly_aligned, by = "date")
cat("after fuel:", nrow(df_weekly), "\n")

df_weekly <- df_weekly %>% dplyr::left_join(df_cpi_weekly, by = "date")
cat("after cpi:", nrow(df_weekly), "\n")

df_weekly <- df_weekly %>% dplyr::left_join(df_api_weekly, by = "date")
cat("after api:", nrow(df_weekly), "\n")

df_weekly <- df_weekly %>% dplyr::left_join(df_fert_weekly, by = "date")
cat("after fert:", nrow(df_weekly), "\n")

df_weekly <- df_weekly %>% dplyr::left_join(df_sppi_weekly, by = "date")
cat("after sppi:", nrow(df_weekly), "\n")

df_weekly <- df_weekly %>% dplyr::arrange(date)

# ── Build df_weekly ───────────────────────────────────────────────────────────
df_weekly <- df_price_weekly %>%
  dplyr::left_join(df_fuel_weekly_aligned, by = "date") %>%
  dplyr::left_join(df_cpi_weekly,          by = "date") %>%
  dplyr::left_join(df_api_weekly,          by = "date") %>%
  dplyr::left_join(df_fert_weekly,         by = "date") %>%
  dplyr::left_join(df_sppi_weekly,         by = "date") %>%
  dplyr::arrange(date)

# Replace stopifnot with informative check
cat("df_weekly rows:", nrow(df_weekly), "\n")
cat("weekly_index rows:", nrow(weekly_index), "\n")
cat("df_price_weekly rows:", nrow(df_price_weekly), "\n")

if (nrow(df_weekly) != nrow(weekly_index)) {
  warning(paste("Row mismatch: df_weekly =", nrow(df_weekly), 
                "vs weekly_index =", nrow(weekly_index)))
}
```
```{r}
library(tseries)   # adf.test
library(vars)      # VAR, VARselect, causality
library(lmtest)    # grangertest
library(dplyr)
library(tidyr)
library(purrr)

predictors <- c("fuel_price", "cpi", "api", "fertiliser", "sppi")
y_var      <- "price"
var_list   <- c(y_var, predictors)
```
```{r}
run_adf <- function(df, var_names, freq_label) {   # renamed vars → var_names
  purrr::map_dfr(var_names, function(v) {           # updated here too
    series <- df[[v]]
    series <- series[!is.na(series)]
    
    adf_level <- tryCatch(
      tseries::adf.test(series),
      error = function(e) list(p.value = NA)
    )
    
    adf_diff <- tryCatch(
      tseries::adf.test(diff(series)),
      error = function(e) list(p.value = NA)
    )
    
    tibble::tibble(
      frequency             = freq_label,
      variable              = v,
      adf_level_p           = round(adf_level$p.value, 4),
      adf_diff_p            = round(adf_diff$p.value, 4),
      stationary_at_level   = adf_level$p.value < 0.05,
      stationary_after_diff = adf_diff$p.value < 0.05
    )
  })
}

# Then call with var_list instead
adf_weekly    <- run_adf(df_weekly,    var_list, "weekly")
adf_monthly   <- run_adf(df_monthly,   var_list, "monthly")
adf_quarterly <- run_adf(df_quarterly, var_list, "quarterly")

adf_results <- dplyr::bind_rows(adf_weekly, adf_monthly, adf_quarterly)

cat("\n══ STEP 1: ADF Stationarity Tests ══\n")
print(adf_results)
```
```{r}
# Use second difference for weekly fuel_price only
df_weekly <- df_weekly %>%
  dplyr::arrange(date) %>%
  dplyr::mutate(
    fuel_price_d2 = fuel_price - dplyr::lag(fuel_price, 1) - 
                   dplyr::lag(fuel_price - dplyr::lag(fuel_price), 1)
  )

# Confirm it's now stationary
fuel_d2 <- df_weekly$fuel_price_d2[!is.na(df_weekly$fuel_price_d2)]
tseries::adf.test(fuel_d2)
```
```{r}
# Weekly: swap fuel_price for fuel_price_d2
predictors_weekly    <- c("fuel_price_d2", "cpi", "api", "fertiliser", "sppi")

# Monthly and quarterly: fuel_price first difference is fine
predictors_monthly    <- c("fuel_price", "cpi", "api", "fertiliser", "sppi")
predictors_quarterly  <- c("fuel_price", "cpi", "api", "fertiliser", "sppi")
```
```{r}
# ── Step 2: Lag Selection ─────────────────────────────────────────────────────
run_lag_selection <- function(df, y, xs, freq_label, max_lag = 8) {
  
  xs_avail <- xs[sapply(xs, function(x) sum(!is.na(df[[x]])) > max_lag * 2)]
  
  d <- df %>%
    dplyr::arrange(date) %>%
    dplyr::mutate(
      dplyr::across(dplyr::all_of(c(y, xs_avail)), ~ .x - dplyr::lag(.x), .names = "d_{.col}")
    ) %>%
    tidyr::drop_na()
  
  var_df <- d %>% dplyr::select(dplyr::starts_with("d_"))
  
  sel <- vars::VARselect(var_df, lag.max = max_lag, type = "const")
  
  cat("\n══ STEP 2: Lag Selection —", freq_label, "══\n")
  cat("AIC suggests lag:", sel$selection["AIC(n)"], "\n")
  cat("BIC suggests lag:", sel$selection["SC(n)"],  "\n")
  print(sel$criteria)
  
  as.integer(sel$selection["SC(n)"])
}

# ── Step 3: Pairwise Granger ──────────────────────────────────────────────────
run_pairwise <- function(df, y, xs, lags, freq_label) {
  
  cat("\n══ STEP 3: Pairwise Granger —", freq_label, "(lag =", lags, ") ══\n")
  cat("H0: predictor does NOT Granger-cause", y, "\n\n")
  
  results <- purrr::map_dfr(xs, function(x) {
    
    d <- df %>%
      dplyr::arrange(date) %>%
      dplyr::mutate(
        dy = .data[[y]] - dplyr::lag(.data[[y]]),
        dx = .data[[x]] - dplyr::lag(.data[[x]])
      ) %>%
      tidyr::drop_na(dy, dx)
    
    if (nrow(d) < lags * 3) {
      return(tibble::tibble(frequency = freq_label, predictor = x,
                            lag = lags, f_stat = NA, p_value = NA,
                            significant = "insufficient data"))
    }
    
    gt <- tryCatch(
      lmtest::grangertest(dy ~ dx, order = lags, data = d),
      error = function(e) NULL
    )
    
    if (is.null(gt)) {
      return(tibble::tibble(frequency = freq_label, predictor = x,
                            lag = lags, f_stat = NA, p_value = NA,
                            significant = "error"))
    }
    
    p <- gt$`Pr(>F)`[2]
    f <- gt$F[2]
    sig <- if (p < 0.01) "*** (1%)" else if (p < 0.05) "**  (5%)" else if (p < 0.10) "*   (10%)" else "not significant"
    
    tibble::tibble(
      frequency   = freq_label,
      predictor   = x,
      lag         = lags,
      f_stat      = round(f, 3),
      p_value     = round(p, 4),
      significant = sig
    )
  })
  
  print(results)
  results
}

# ── Step 4: Joint Granger ─────────────────────────────────────────────────────
run_joint <- function(df, y, xs, lags, freq_label) {
  
  cat("\n══ STEP 4: Joint Granger —", freq_label, "(lag =", lags, ") ══\n")
  cat("H0: ALL predictors jointly do NOT Granger-cause", y, "\n\n")
  
  xs_avail <- xs[sapply(xs, function(x) sum(!is.na(df[[x]])) > lags * 2)]
  if (length(xs_avail) == 0) { cat("No predictors available.\n"); return(NULL) }
  
  d <- df %>%
    dplyr::arrange(date) %>%
    dplyr::mutate(dplyr::across(
      dplyr::all_of(c(y, xs_avail)),
      ~ .x - dplyr::lag(.x), .names = "d_{.col}"
    )) %>%
    tidyr::drop_na()
  
  var_df <- d %>% dplyr::select(dplyr::all_of(paste0("d_", c(y, xs_avail))))
  
  fit <- tryCatch(
    vars::VAR(var_df, p = lags, type = "const"),
    error = function(e) { cat("VAR failed:", e$message, "\n"); NULL }
  )
  if (is.null(fit)) return(NULL)
  
  caus <- tryCatch(
    vars::causality(fit, cause = paste0("d_", xs_avail)),
    error = function(e) { cat("Causality test failed:", e$message, "\n"); NULL }
  )
  if (is.null(caus)) return(NULL)
  
  p_val   <- unname(as.vector(caus$Granger$p.value))[1]
  chi_val <- unname(as.vector(caus$Granger$statistic))[1]
  df_val  <- unname(as.vector(caus$Granger$parameter))[1]
  sig     <- if (p_val < 0.01) "*** (1%)" else if (p_val < 0.05) "**  (5%)" else if (p_val < 0.10) "*   (10%)" else "not significant"
  
  result <- tibble::tibble(
    frequency       = freq_label,
    test            = "Joint (all predictors → price)",
    predictors_used = paste(xs_avail, collapse = ", "),
    lag             = lags,
    chi_sq          = round(chi_val, 3),
    df_test         = df_val,
    p_value         = round(p_val, 4),
    significant     = sig
  )
  print(result)
  result
}

# ── Step 5: Diagnostics ───────────────────────────────────────────────────────
run_diagnostics <- function(df, y, xs, lags, freq_label) {
  
  cat("\n══ STEP 5: Residual Diagnostics —", freq_label, "══\n")
  
  xs_avail <- xs[sapply(xs, function(x) sum(!is.na(df[[x]])) > lags * 2)]
  
  d <- df %>%
    dplyr::arrange(date) %>%
    dplyr::mutate(
      dplyr::across(
        dplyr::all_of(c(y, xs_avail)),
        ~ .x - dplyr::lag(.x),
        .names = "d_{.col}"
      )
    ) %>%
    tidyr::drop_na()
  
  var_df <- d %>%
    dplyr::select(dplyr::all_of(paste0("d_", c(y, xs_avail))))
  
  fit <- tryCatch(
    vars::VAR(var_df, p = lags, type = "const"),
    error = function(e) { cat("VAR failed:", e$message, "\n"); NULL }
  )
  if (is.null(fit)) { cat("VAR failed, skipping diagnostics.\n"); return(NULL) }
  
  # Serial correlation test
  tryCatch({
    serial <- vars::serial.test(fit, lags.pt = lags + 2, type = "PT.asymptotic")
    p_serial <- unname(as.vector(serial$serial$p.value))[1]
    cat("Portmanteau test p-value:", round(p_serial, 4),
        ifelse(p_serial > 0.05,
               "✓ No autocorrelation in residuals",
               "✗ Autocorrelation detected — consider more lags"), "\n")
  }, error = function(e) {
    cat("Serial test failed (likely singular matrix — common with small quarterly samples):", 
        e$message, "\n")
  })
  
  # Normality test
  tryCatch({
    norm <- vars::normality.test(fit)
    p_norm <- unname(as.vector(norm$jb.mul$JB$p.value))[1]
    cat("Normality test (JB) p-value:", round(p_norm, 4), "\n")
  }, error = function(e) {
    cat("Normality test failed:", e$message, "\n")
  })
  
  invisible(fit)
}
```

```{r}
# ── Step 2: Lag Selection ────────────────────────────────────────────────────
lag_weekly    <- run_lag_selection(df_weekly,    y_var, predictors_weekly,    "Weekly",    max_lag = 8)
lag_monthly   <- run_lag_selection(df_monthly,   y_var, predictors_monthly,   "Monthly",   max_lag = 6)
lag_quarterly <- run_lag_selection(df_quarterly, y_var, predictors_quarterly, "Quarterly", max_lag = 4)
```
```{r}
# ── Step 3: Pairwise ─────────────────────────────────────────────────────────
pairwise_weekly    <- run_pairwise(df_weekly,    y_var, predictors_weekly,    lag_weekly,    "Weekly")
pairwise_monthly   <- run_pairwise(df_monthly,   y_var, predictors_monthly,   lag_monthly,   "Monthly")
pairwise_quarterly <- run_pairwise(df_quarterly, y_var, predictors_quarterly, lag_quarterly, "Quarterly")
```
```{r}
# ── Step 4: Joint ────────────────────────────────────────────────────────────
joint_weekly    <- run_joint(df_weekly,    y_var, predictors_weekly,    lag_weekly,    "Weekly")
joint_monthly   <- run_joint(df_monthly,   y_var, predictors_monthly,   lag_monthly,   "Monthly")
joint_quarterly <- run_joint(df_quarterly, y_var, predictors_quarterly, lag_quarterly, "Quarterly")
```
```{r}
run_diagnostics(df_weekly,    y_var, predictors_weekly,    lag_weekly,    "Weekly")
run_diagnostics(df_monthly,   y_var, predictors_monthly,   lag_monthly,   "Monthly")
run_diagnostics(df_quarterly, y_var, predictors_quarterly, lag_quarterly, "Quarterly")
```
```{r}
pairwise_all <- dplyr::bind_rows(pairwise_weekly, pairwise_monthly, pairwise_quarterly)
joint_all    <- dplyr::bind_rows(joint_weekly, joint_monthly, joint_quarterly)

cat("\n\n══ FINAL: Pairwise Granger Results ══\n")
print(pairwise_all)

cat("\n══ FINAL: Joint Granger Results ══\n")
print(joint_all)
```
```{r}
# ── Fixes ─────────────────────────────────────────────────────────────────────

# Weekly: drop sppi (LOCF from quarterly → all-zero after diff → singular)
predictors_weekly_joint <- c("fuel_price_d2", "cpi", "api", "fertiliser")

# Monthly: use AIC lag (3) instead of BIC (1) to fix autocorrelation
lag_monthly <- 3

# Quarterly: cap at lag 2 (45 rows can't support lag 4 with 6 variables)
lag_quarterly <- 2

# ── Fix run_joint: extract p.value as scalar ──────────────────────────────────
run_joint_fixed <- function(df, y, xs, lags, freq_label) {
  
  cat("\n══ STEP 4: Joint Granger —", freq_label, "(lag =", lags, ") ══\n")
  cat("H0: ALL predictors jointly do NOT Granger-cause", y, "\n\n")
  
  xs_avail <- xs[sapply(xs, function(x) sum(!is.na(df[[x]])) > lags * 2)]
  
  if (length(xs_avail) == 0) {
    cat("No predictors available.\n")
    return(NULL)
  }
  
  d <- df %>%
    dplyr::arrange(date) %>%
    dplyr::mutate(
      dplyr::across(
        dplyr::all_of(c(y, xs_avail)),
        ~ .x - dplyr::lag(.x),
        .names = "d_{.col}"
      )
    ) %>%
    tidyr::drop_na()
  
  var_df <- d %>%
    dplyr::select(dplyr::all_of(paste0("d_", c(y, xs_avail))))
  
  mat <- as.matrix(var_df)
  if (rcond(cov(mat)) < 1e-15) {
    cat("Matrix is singular — likely perfect collinearity among predictors at this frequency.\n")
    return(NULL)
  }
  
  fit <- tryCatch(
    vars::VAR(var_df, p = lags, type = "const"),
    error = function(e) { cat("VAR failed:", e$message, "\n"); NULL }
  )
  if (is.null(fit)) return(NULL)
  
  caus <- tryCatch(
    vars::causality(fit, cause = paste0("d_", xs_avail)),
    error = function(e) { cat("Causality test failed:", e$message, "\n"); NULL }
  )
  if (is.null(caus)) return(NULL)
  
  p_val   <- unname(as.vector(caus$Granger$p.value))[1]
  chi_val <- unname(as.vector(caus$Granger$statistic))[1]
  df_val  <- unname(as.vector(caus$Granger$parameter))[1]
  sig     <- if (p_val < 0.01) "*** (1%)" else if (p_val < 0.05) "**  (5%)" else if (p_val < 0.10) "*   (10%)" else "not significant"
  
  result <- tibble::tibble(
    frequency       = freq_label,
    test            = "Joint (all predictors → price)",
    predictors_used = paste(xs_avail, collapse = ", "),
    lag             = lags,
    chi_sq          = round(chi_val, 3),
    df_test         = df_val,
    p_value         = round(p_val, 4),
    significant     = sig
  )
  
  print(result)
  result
}

```
```{r}
# ── Re-run Step 3 with corrected lags ────────────────────────────────────────
pairwise_monthly   <- run_pairwise(df_monthly,   y_var, predictors_monthly,   lag_monthly,   "Monthly")
pairwise_quarterly <- run_pairwise(df_quarterly, y_var, predictors_quarterly, lag_quarterly, "Quarterly")
```
```{r}
# ── Re-run Step 4 with fixes ──────────────────────────────────────────────────
joint_weekly    <- run_joint_fixed(df_weekly,    y_var, predictors_weekly_joint, lag_weekly,    "Weekly")
joint_monthly   <- run_joint_fixed(df_monthly,   y_var, predictors_monthly,      lag_monthly,   "Monthly")
joint_quarterly <- run_joint_fixed(df_quarterly, y_var, predictors_quarterly,    lag_quarterly, "Quarterly")
```
```{r}
# ── Re-run Step 5 diagnostics ─────────────────────────────────────────────────
run_diagnostics(df_weekly,    y_var, predictors_weekly_joint, lag_weekly,    "Weekly")
run_diagnostics(df_monthly,   y_var, predictors_monthly,      lag_monthly,   "Monthly")
run_diagnostics(df_quarterly, y_var, predictors_quarterly,    lag_quarterly, "Quarterly")
```
```{r}
# ── Final Summary ──────────────────────────────────────────────────────────────
pairwise_all <- dplyr::bind_rows(pairwise_weekly, pairwise_monthly, pairwise_quarterly)
joint_all    <- dplyr::bind_rows(joint_weekly, joint_monthly, joint_quarterly)

cat("\n\n══ FINAL: Pairwise Granger Results ══\n")
print(pairwise_all)

cat("\n══ FINAL: Joint Granger Results ══\n")
print(joint_all)
```
```{r}
# ── Clean summary for reporting ───────────────────────────────────────────────
summary_table <- pairwise_all %>%
  dplyr::mutate(
    result = ifelse(p_value < 0.01, "Strong (1%)",
             ifelse(p_value < 0.05, "Moderate (5%)",
             ifelse(p_value < 0.10, "Weak (10%)", "None")))
  ) %>%
  dplyr::select(frequency, predictor, lag, p_value, result) %>%
  tidyr::pivot_wider(
    names_from  = frequency,
    values_from = c(p_value, result)
  )

cat("\n══ GRANGER CAUSALITY SUMMARY (X → fruit & veg price) ══\n")
print(summary_table)

# ── Key findings ──────────────────────────────────────────────────────────────
cat("
KEY FINDINGS:
- CPI → price: Significant at monthly (p=0.038) and quarterly (p=0.004) — most robust result
- Fuel price → price: Marginal at weekly only (p=0.094) — short-run effect only
- SPPI → price: Marginal at monthly only (p=0.100) — weak signal
- API, Fertiliser → price: No significant Granger causality at any frequency
- Joint tests: Non-significant — CPI signal diluted when all predictors included together
- Weekly joint VAR: Inconclusive (matrix singularity due to SPPI collinearity)
")
```
**Granger Causality Analysis: Determinants of UK Fruit & Vegetable Price Movements**

---

**Overview**

This section tests whether macroeconomic and agricultural cost variables Granger-cause movements in UK retail fruit and vegetable prices. Granger causality tests whether the past values of a predictor variable contain statistically significant information about the future values of a target variable, beyond what the target's own history already provides. A significant result does not imply structural causation — it indicates that the predictor has *predictive* content for the outcome. Tests were conducted at three temporal resolutions — weekly, monthly, and quarterly — and in two modes: pairwise (each predictor tested independently) and joint (all predictors tested simultaneously via a Vector Autoregression). The predictors examined are fuel price, Consumer Price Index (CPI), Agricultural Price Index (API), fertiliser price, and the Services Producer Price Index (SPPI).

---

**Methodology**

Prior to Granger testing, all series were assessed for stationarity using the Augmented Dickey-Fuller (ADF) test at both levels and first differences. Most series were found to be integrated of order one (I(1)) — non-stationary in levels but stationary after first differencing — satisfying the standard precondition for Granger testing in a VAR framework. One exception was fuel price at the weekly frequency, which was found to be I(2), requiring a second difference to achieve stationarity. Accordingly, the second-differenced series (fuel_price_d2) was substituted in weekly-level tests only. Fertiliser data was unavailable prior to 2017, resulting in partial coverage at monthly and quarterly frequencies.

Optimal lag lengths were selected using VARselect applied to first-differenced data, with the Bayesian Information Criterion (BIC) used as the primary selection criterion given its stronger penalty for over-parameterisation. BIC selected lag 1 at both weekly and monthly frequencies, and lag 4 at quarterly. However, the monthly lag was overridden to 3 (per AIC) after the BIC-selected model produced serially correlated residuals (Portmanteau p = 0.042). The quarterly lag was capped at 2 due to sample size constraints — with only 45 observations, a lag-4 VAR across six variables is severely over-parameterised and produces a singular coefficient matrix.

---

**Pairwise Granger Causality Results**

CPI is the dominant predictor and the only variable to achieve significance across multiple frequencies. It reaches significance at the monthly level (F = 2.90, p = 0.038) and strong significance at the quarterly level (F = 6.63, p = 0.004). This pattern suggests that inflationary pressures propagate into retail food prices with a lag of one to two quarters, consistent with the broader macroeconomic literature on price transmission. The quarterly result in particular is robust — a p-value of 0.004 at lag 2 provides strong evidence that CPI movements reliably precede fruit and vegetable price changes at that horizon.

Fuel price shows a marginally significant effect at the weekly frequency only (F = 3.08, p = 0.094), using the second-differenced series. This signal does not persist at monthly or quarterly frequencies, suggesting that any fuel price transmission into food prices is confined to very short-run, week-to-week fluctuations and dissipates quickly. This may reflect the role of fuel costs in immediate distribution and logistics, which are absorbed or passed through within days rather than propagating into longer-run price setting.

SPPI shows a borderline monthly effect (F = 2.13, p = 0.100) that warrants cautious interpretation. It may reflect indirect cost pressures through logistics and distribution services, but the evidence sits at the threshold of conventional significance and should not be overstated.

Neither the Agricultural Price Index nor fertiliser prices exhibit significant Granger causality at any frequency or lag specification. This is a notable null result. Despite the intuitive prior that upstream input costs should drive retail food prices, there is no statistical evidence of a predictive relationship in this dataset over the sample period. This may reflect the complexity of agricultural supply chains, producer hedging behaviour, or the dominance of imported produce in UK retail, where domestic input cost structures are less directly relevant.

---

**Joint Granger Causality Results**

Neither the monthly (χ² = 1.18, p = 0.284) nor the quarterly (χ² = 1.02, p = 0.429) joint tests reach significance. This apparent contradiction with the pairwise CPI results is a well-documented phenomenon in Granger causality analysis: joint VAR tests carry substantially more degrees of freedom — 15 for monthly and 10 for quarterly — which reduces statistical power when only one predictor in the system is genuinely informative while the rest are noise. The non-significance of the joint test does not invalidate the pairwise CPI finding; it reflects the diluting effect of jointly testing multiple null predictors alongside a real signal.

The weekly joint VAR could not be estimated due to matrix singularity. SPPI, filled via last-observation-carried-forward from quarterly data, produces a constant series within each quarter and collapses to zero after first differencing, rendering the covariance matrix rank-deficient. This predictor was excluded from the weekly joint test, but the remaining variables still produced a singular matrix at lag 1, likely due to near-perfect collinearity among the short differenced series. The weekly frequency is therefore best analysed through pairwise tests only.

---

**Residual Diagnostics**

The monthly VAR at lag 3 passes the Portmanteau test for serial correlation, confirming that the override from BIC's lag-1 selection was appropriate. Non-normality of residuals (Jarque-Bera p ≈ 0) is observed across frequencies but is expected given the presence of price spikes and supply disruptions in the sample period — including COVID-19 pandemic effects and post-Brexit trade adjustments — and does not invalidate the Granger test results given the large sample sizes available at weekly and monthly frequencies.

---

**Limitations**

Several limitations should be noted. First, Granger causality is a test of predictive precedence, not structural causation — significant results indicate timing relationships but not underlying mechanisms. Second, the quarterly analysis suffers from low statistical power (n = 45) and stationarity test results at this frequency should be interpreted cautiously, as the ADF test has well-known size distortions in small samples. Third, the fertiliser series covers only 2017 onward, limiting the power of tests involving this variable. Fourth, the analysis assumes linear, time-invariant relationships — structural breaks associated with Brexit and the COVID-19 pandemic may violate this assumption, and rolling-window Granger tests could be a productive extension. Finally, the LOCF treatment of quarterly SPPI data introduces artificial constancy at weekly and monthly frequencies; a quarterly-only analysis of SPPI's relationship with price is more methodologically defensible.

---

**Summary**

CPI is the most robust leading indicator of UK fruit and vegetable price movements, with significant pairwise Granger causality at both monthly (p = 0.038) and quarterly (p = 0.004) frequencies. Fuel price shows a weak, transient short-run effect at the weekly level only, while SPPI shows a borderline monthly signal. API and fertiliser prices exhibit no significant predictive relationship with food prices at any frequency — a notable null result given the economic prior. Joint VAR tests are non-significant across all frequencies, reflecting reduced power in the presence of multiple null predictors, and the weekly joint test is inconclusive due to matrix singularity. Taken together, the results point to demand-side inflation as the primary macroeconomic driver of fruit and vegetable price dynamics in the UK, with supply-side cost factors playing a less discernible role within this analytical framework.