---
title: "Final Project: GARCH Model"
author: "Minhao Zhang"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, fig.width = 12, fig.height = 6)
```

# 0 Load Libraries and Data

```{r load_libraries}
library(arrow)
library(rugarch)
library(TSA)
library(ggplot2)
library(gridExtra)
library(tseries)
library(moments)
library(knitr)
library(FinTS)
library(dplyr)
library(tidyr)
```

```{r define_colors}
col_blue <- "#2E86AB"
col_red <- "#F24236"
col_green <- "#06A77D"
col_purple <- "#9B59B6"
```

```{r load_data}
df_raw <- read_parquet("../../data/processed/merged_features_weekly.parquet")

commodities <- c(
  "Beetroot", "Bulb Onions (Yellow)", "Cabbage", "Carrots",
  "Cucumbers", "Curly kale", "Lettuce", "Tomatoes (Round)"
)

df <- df_raw %>%
  filter(commodity %in% commodities) %>%
  arrange(commodity, date)

compute_log_returns <- function(prices) {
  ret <- diff(log(prices)) * 100
  return(ret)
}

test_weeks <- 52
commodity_data <- list()

for (comm in commodities) {
  sub <- df %>% filter(commodity == comm)
  prices <- sub$target_price

  # Calculate returns for the whole series
  ret_full <- compute_log_returns(prices)
  dates_full <- sub$date[-1]

  n <- length(ret_full)
  if (n > test_weeks) {
    # Split into train (everything except last 52 weeks) and test (last 52 weeks)
    train_idx <- 1:(n - test_weeks)
    test_idx <- (n - test_weeks + 1):n

    ret_train <- ret_full[train_idx]
    dates_train <- dates_full[train_idx]

    ret_test <- ret_full[test_idx]
    dates_test <- dates_full[test_idx]

    commodity_data[[comm]] <- list(
      prices = prices[c(train_idx, n)], # Approximate
      returns = ret_train, # "returns" used by EDA is now TRAIN returns
      dates = dates_train,
      test_returns = ret_test,
      test_dates = dates_test
    )
  } else {
    warning(paste("Not enough data for", comm, "to split 52 weeks."))
  }
}
```

# 1 Exploratory Data Analysis

We use **Carrots** as a representative commodity for the detailed EDA.

## 1.1 Time Series Visualization

```{r ts_plot}
eda_comm <- "Carrots"
eda <- commodity_data[[eda_comm]]

df_returns <- data.frame(Date = eda$dates, Return = eda$returns)

ggplot(df_returns, aes(x = Date, y = Return)) +
  geom_line(color = col_blue, size = 0.5) +
  theme_minimal() +
  labs(
    title = paste0(eda_comm, " Weekly Log Returns (%)"),
    x = "Date", y = "Return (%)"
  ) +
  theme(plot.title = element_text(hjust = 0.5, size = 14, face = "bold"))
```

## 1.2 Distribution Analysis

```{r distribution, fig.height=5}
p1 <- ggplot(df_returns, aes(x = Return)) +
  geom_histogram(bins = 60, fill = col_blue, color = "white", alpha = 0.7) +
  geom_vline(
    xintercept = mean(eda$returns), color = col_red,
    linetype = "dashed", size = 1
  ) +
  theme_minimal() +
  labs(
    title = "Return Distribution",
    subtitle = paste0("Mean = ", round(mean(eda$returns), 3), "%"),
    x = "Return (%)", y = "Frequency"
  ) +
  theme(
    plot.title = element_text(hjust = 0.5, size = 12, face = "bold"),
    plot.subtitle = element_text(hjust = 0.5, size = 10)
  )

p2 <- ggplot(df_returns, aes(sample = Return)) +
  stat_qq(color = col_blue) +
  stat_qq_line(color = col_red, size = 1) +
  theme_minimal() +
  labs(
    title = "Q-Q Plot (Normality Check)",
    x = "Theoretical Quantiles", y = "Sample Quantiles"
  ) +
  theme(plot.title = element_text(hjust = 0.5, size = 12, face = "bold"))

grid.arrange(p1, p2, ncol = 2)
```

## 1.3 Descriptive Statistics

```{r desc_stats}
ret <- eda$returns
summary(ret)
cat("Skewness:", round(skewness(ret), 4), "\n")
cat("Kurtosis:", round(kurtosis(ret), 4), "\n")
```

## 1.4 Autocorrelation of Returns

```{r acf_returns}
par(mfrow = c(2, 1))
acf(ret, main = paste0("ACF: ", eda_comm, " Returns"), col = col_blue, lwd = 2, lag.max = 40)
pacf(ret, main = paste0("PACF: ", eda_comm, " Returns"), col = col_blue, lwd = 2, lag.max = 40)
par(mfrow = c(1, 1))
```

## 1.5 Ljung-Box Test on Returns

```{r ljung_box_returns}
lags_test <- c(5, 10, 15, 20, 25, 30)
lb_results_returns <- data.frame(Lag = numeric(), Statistic = numeric(), PValue = numeric())

for (lag in lags_test) {
  lb_test <- Box.test(ret, lag = lag, type = "Ljung-Box")
  lb_results_returns <- rbind(lb_results_returns, data.frame(
    Lag = lag,
    Statistic = round(unname(lb_test$statistic), 4),
    PValue = round(unname(lb_test$p.value), 6)
  ))
}
kable(lb_results_returns, caption = "Ljung-Box Test on Returns")
```

## 1.6 Squared Returns (Volatility Proxy)

```{r squared_returns}
ret_sq <- ret^2
df_squared <- data.frame(Date = eda$dates, SquaredReturn = ret_sq)

ggplot(df_squared, aes(x = Date, y = SquaredReturn)) +
  geom_line(color = col_red, size = 0.5) +
  theme_minimal() +
  labs(
    title = "Squared Returns (Volatility Clustering)",
    x = "Date", y = "Squared Return"
  ) +
  theme(plot.title = element_text(hjust = 0.5, size = 14, face = "bold"))
```

## 1.7 ACF/PACF of Squared Returns

```{r acf_squared}
par(mfrow = c(2, 1))
acf(ret_sq, main = "ACF: Squared Returns", col = col_red, lwd = 2, lag.max = 40)
pacf(ret_sq, main = "PACF: Squared Returns", col = col_red, lwd = 2, lag.max = 40)
par(mfrow = c(1, 1))
```

## 1.8 Test for Heteroscedasticity

### 1.8.1 McLeod-Li Test

```{r mcleod_li}
mcleod_li_results <- data.frame(Lag = numeric(), PValue = numeric())
for (lag in lags_test) {
  ml_test <- McLeod.Li.test(y = ret, gof.lag = lag, plot = FALSE)
  mcleod_li_results <- rbind(mcleod_li_results, data.frame(
    Lag = lag,
    PValue = round(ml_test$p.values[lag], 6)
  ))
}
kable(mcleod_li_results, caption = "McLeod-Li Test for Heteroscedasticity")
```

### 1.8.2 Ljung-Box Test on Squared Returns

```{r ljung_box_squared}
lb_results_squared <- data.frame(Lag = numeric(), Statistic = numeric(), PValue = numeric())

for (lag in lags_test) {
  lb_test_sq <- Box.test(ret_sq, lag = lag, type = "Ljung-Box")
  lb_results_squared <- rbind(lb_results_squared, data.frame(
    Lag = lag,
    Statistic = round(unname(lb_test_sq$statistic), 4),
    PValue = round(unname(lb_test_sq$p.value), 6)
  ))
}
kable(lb_results_squared, caption = "Ljung-Box Test on Squared Returns")
```

### 1.8.3 ARCH-LM Test

```{r arch_lm_test}
arch_test <- ArchTest(ret, lags = 12)
cat("ARCH-LM Test:\n")
cat("  Chi-squared:", round(arch_test$statistic, 4), "\n")
cat("  p-value:", format(arch_test$p.value, scientific = TRUE), "\n")
```

### 1.8.4 Comparison Plot

```{r hetero_comparison}
comparison <- data.frame(
  Lag = lags_test,
  McLeod_Li = mcleod_li_results$PValue,
  LB_Squared = lb_results_squared$PValue
)

ggplot(comparison, aes(x = Lag)) +
  geom_line(aes(y = McLeod_Li, color = "McLeod-Li"), size = 1.2) +
  geom_point(aes(y = McLeod_Li, color = "McLeod-Li"), size = 3) +
  geom_line(aes(y = LB_Squared, color = "LB on Squared"),
    size = 1.2, linetype = "dashed"
  ) +
  geom_point(aes(y = LB_Squared, color = "LB on Squared"), size = 3) +
  geom_hline(yintercept = 0.05, linetype = "dotted", color = "gray30") +
  scale_color_manual(values = c(
    "McLeod-Li" = col_green,
    "LB on Squared" = col_red
  )) +
  theme_minimal() +
  labs(
    title = "P-Values: McLeod-Li vs Ljung-Box on Squared Returns",
    x = "Lag", y = "p-value",
    color = "Test"
  ) +
  theme(
    plot.title = element_text(hjust = 0.5, size = 14, face = "bold"),
    legend.position = "top"
  )
```

## 1.9 Rolling Volatility

```{r rolling_vol}
window_size <- 8
rolling_vol <- zoo::rollapply(ret,
  width = window_size, FUN = sd,
  fill = NA, align = "right"
)

df_rolling <- data.frame(
  Date = eda$dates,
  Rolling_Vol = as.numeric(rolling_vol)
)
df_rolling <- na.omit(df_rolling)

ggplot(df_rolling, aes(x = Date, y = Rolling_Vol)) +
  geom_line(color = col_red, size = 1) +
  theme_minimal() +
  labs(
    title = "8-Week Rolling Volatility",
    x = "Date", y = "Volatility (%)"
  ) +
  theme(plot.title = element_text(hjust = 0.5, size = 14, face = "bold"))
```

# 2 GARCH Model

## 2.1 GARCH Order Selection (Per Commodity)

```{r garch_order_selection}
garch_orders <- expand.grid(p = 1:3, q = 1:3)
garch_results <- list()

for (comm in commodities) {
  ret_c <- commodity_data[[comm]]$returns
  garch_aic <- numeric(nrow(garch_orders))
  garch_bic <- numeric(nrow(garch_orders))

  for (i in 1:nrow(garch_orders)) {
    spec <- ugarchspec(
      variance.model = list(
        model = "sGARCH",
        garchOrder = c(garch_orders$p[i], garch_orders$q[i])
      ),
      mean.model = list(armaOrder = c(0, 0), include.mean = TRUE),
      distribution.model = "norm"
    )
    fit <- tryCatch(
      ugarchfit(spec, ret_c, solver = "hybrid"),
      error = function(e) NULL
    )
    if (!is.null(fit)) {
      garch_aic[i] <- infocriteria(fit)[1]
      garch_bic[i] <- infocriteria(fit)[2]
    } else {
      garch_aic[i] <- NA
      garch_bic[i] <- NA
    }
  }

  best_idx <- which.min(garch_bic)
  garch_results[[comm]] <- list(
    aic = garch_aic,
    bic = garch_bic,
    best_p = garch_orders$p[best_idx],
    best_q = garch_orders$q[best_idx]
  )
}

order_summary <- data.frame(
  Commodity = commodities,
  Best_p = sapply(commodities, function(c) garch_results[[c]]$best_p),
  Best_q = sapply(commodities, function(c) garch_results[[c]]$best_q),
  Best_BIC = sapply(commodities, function(c) round(min(garch_results[[c]]$bic, na.rm = TRUE), 4))
)
order_summary$Model <- paste0("GARCH(", order_summary$Best_p, ",", order_summary$Best_q, ")")
kable(order_summary[, c("Commodity", "Model", "Best_BIC")],
  caption = "Best GARCH(p,q) Order by BIC", row.names = FALSE
)
```

## 2.2 GARCH Order Selection Detail (Example)

```{r garch_order_detail}
garch_ic <- data.frame(
  Model = paste0("GARCH(", garch_orders$p, ",", garch_orders$q, ")"),
  p = garch_orders$p,
  q = garch_orders$q,
  AIC = round(garch_results[[eda_comm]]$aic, 4),
  BIC = round(garch_results[[eda_comm]]$bic, 4)
)
kable(garch_ic[, c("Model", "AIC", "BIC")],
  caption = paste0("GARCH(p,q) Information Criteria — ", eda_comm)
)
```

## 2.3 Fit GARCH Models

```{r fit_garch}
garch_fits <- list()

for (comm in commodities) {
  ret_c <- commodity_data[[comm]]$returns
  best_p <- garch_results[[comm]]$best_p
  best_q <- garch_results[[comm]]$best_q

  spec <- ugarchspec(
    variance.model = list(model = "sGARCH", garchOrder = c(best_p, best_q)),
    mean.model = list(armaOrder = c(0, 0), include.mean = TRUE),
    distribution.model = "norm"
  )
  garch_fits[[comm]] <- ugarchfit(spec, ret_c, solver = "hybrid")
}

cat("Example: GARCH model for", eda_comm, "\n\n")
garch_fits[[eda_comm]]
```

## 2.4 GARCH Residual Analysis

```{r garch_residuals, fig.height=8}
garch_std_resid <- as.numeric(residuals(garch_fits[[eda_comm]], standardize = TRUE))
garch_sigma <- as.numeric(sigma(garch_fits[[eda_comm]]))

par(mfrow = c(2, 2))
plot(garch_std_resid,
  type = "l", col = col_blue,
  main = paste0(
    "GARCH(", garch_results[[eda_comm]]$best_p, ",",
    garch_results[[eda_comm]]$best_q, ") Standardized Residuals"
  ),
  xlab = "Observation", ylab = "Residual"
)
abline(h = 0, col = "red", lty = 2)

hist(garch_std_resid,
  breaks = 50, col = col_blue, border = "white",
  main = "Residual Distribution", xlab = "Standardized Residual", probability = TRUE
)
curve(dnorm(x), add = TRUE, col = "red", lwd = 2)

acf(garch_std_resid, main = "ACF: Standardized Residuals", col = col_blue, lwd = 2)
acf(garch_std_resid^2, main = "ACF: Squared Std Residuals", col = col_red, lwd = 2)
par(mfrow = c(1, 1))
```

```{r garch_resid_ljungbox}
lb_garch_resid <- data.frame(
  Lag       = lags_test,
  Statistic = numeric(length(lags_test)),
  PValue    = numeric(length(lags_test))
)
for (k in seq_along(lags_test)) {
  lb <- Box.test(garch_std_resid, lag = lags_test[k], type = "Ljung-Box")
  lb_garch_resid$Statistic[k] <- round(lb$statistic, 4)
  lb_garch_resid$PValue[k] <- round(lb$p.value, 4)
}
kable(lb_garch_resid, caption = "Ljung-Box Test on GARCH Standardized Residuals")
```

## 2.5 Conditional Volatility

```{r garch_conditional_vol}
df_garch_vol <- data.frame(Date = commodity_data[[eda_comm]]$dates, Sigma = garch_sigma)

ggplot(df_garch_vol, aes(x = Date, y = Sigma)) +
  geom_line(color = col_red, size = 0.8) +
  theme_minimal() +
  labs(
    title = paste0(
      "GARCH(", garch_results[[eda_comm]]$best_p, ",",
      garch_results[[eda_comm]]$best_q, ") Conditional Volatility — ", eda_comm
    ),
    x = "Date", y = "Conditional SD (%)"
  ) +
  theme(plot.title = element_text(hjust = 0.5, size = 14, face = "bold"))
```

# 3 Volatility Forecasting

## 3.1 Forecasts (52 Weeks Ahead)

```{r garch_forecast}
n_ahead <- 52
forecast_results <- list()

for (comm in commodities) {
  # Forecast using the fitted model (which was fit on training data)
  fcast <- ugarchforecast(garch_fits[[comm]], n.ahead = n_ahead)
  forecast_results[[comm]] <- as.numeric(sigma(fcast))
}

forecast_df <- data.frame(Week = 1:n_ahead)
for (comm in commodities) {
  forecast_df[[comm]] <- round(forecast_results[[comm]], 4)
}

# Show first 10 rows for brevity
kable(head(forecast_df, 10), caption = "GARCH Volatility Forecast (First 10 of 52 Weeks Ahead)")
```

## 3.2 Forecast Visualization

```{r garch_forecast_plot, fig.height=8}
forecast_long <- forecast_df %>%
  pivot_longer(cols = -Week, names_to = "Commodity", values_to = "Volatility")

ggplot(forecast_long, aes(x = Week, y = Volatility, color = Commodity)) +
  geom_line(size = 1) +
  geom_point(size = 2) +
  scale_x_continuous(breaks = seq(1, n_ahead, by = 4)) +
  theme_minimal() +
  labs(
    title = "GARCH Volatility Forecast (52 Weeks Ahead)",
    x = "Weeks Ahead", y = "Forecasted Volatility (SD %)"
  ) +
  theme(
    plot.title = element_text(hjust = 0.5, size = 14, face = "bold"),
    legend.position = "right"
  )
```

## 3.3 Forecast vs Actual Volatility (Test Set Proxy)

We use squared returns of the held-out test set as a proxy for actual volatility to compare with our forecasts.

```{r forecast_vs_actual, fig.height=10}
plots_list <- list()

for (comm in commodities) {
  # Get test data
  test_ret <- commodity_data[[comm]]$test_returns
  test_dates <- commodity_data[[comm]]$test_dates

  # Ensure lengths match (if test set is exactly n_ahead)
  len <- min(length(test_ret), n_ahead)

  df_comp <- data.frame(
    Date = test_dates[1:len],
    Week = 1:len,
    Actual_Sq_Return = test_ret[1:len]^2,
    Forecast_Vol_Sq = forecast_results[[comm]][1:len]^2 # Compare variance (sigma^2) with ret^2
  )

  # Or compare Volatility (Sigma) with Abs Return or just plot together
  # Standard practice: compare sigma^2 with ret^2 or sigma with |ret|

  p <- ggplot(df_comp, aes(x = Week)) +
    geom_line(aes(y = sqrt(Actual_Sq_Return), color = "Actual |Return|"), alpha = 0.5) +
    geom_line(aes(y = sqrt(Forecast_Vol_Sq), color = "Forecast Sigma"), size = 1) +
    labs(
      title = paste(comm, "- Forecast vs Actual Volatility"),
      y = "Volatility / |Return|", x = "Weeks Ahead"
    ) +
    scale_color_manual(values = c("Actual |Return|" = "gray50", "Forecast Sigma" = "red")) +
    theme_minimal()

  plots_list[[comm]] <- p
}

grid.arrange(grobs = plots_list, ncol = 2)
```

## 3.4 Summary Table

```{r garch_summary}
summary_df <- data.frame(
  Commodity = commodities,
  Model = sapply(commodities, function(c) {
    paste0(
      "GARCH(", garch_results[[c]]$best_p, ",",
      garch_results[[c]]$best_q, ")"
    )
  }),
  BIC = sapply(commodities, function(c) round(min(garch_results[[c]]$bic, na.rm = TRUE), 4)),
  LogLikelihood = sapply(commodities, function(c) round(likelihood(garch_fits[[c]]), 2)),
  Alpha_Plus_Beta = sapply(commodities, function(c) {
    coefs <- coef(garch_fits[[c]])
    alpha_sum <- sum(coefs[grep("alpha", names(coefs))])
    beta_sum <- sum(coefs[grep("beta", names(coefs))])
    round(alpha_sum + beta_sum, 4)
  }),
  Vol_Week1 = sapply(commodities, function(c) round(forecast_results[[c]][1], 4)),
  Vol_Week52 = sapply(commodities, function(c) round(forecast_results[[c]][min(52, length(forecast_results[[c]]))], 4))
)
kable(summary_df, caption = "GARCH Model Summary Across Commodities", row.names = FALSE)
```

# 4 Per-Commodity Forecast Metrics

Metrics are computed by comparing the 52-week forecasted conditional volatility (sigma) against absolute test-set returns as a proxy for realized volatility.

```{r garch_metrics}
compute_metrics <- function(actual, predicted) {
  rmse <- sqrt(mean((predicted - actual)^2))
  mae <- mean(abs(predicted - actual))
  # Exclude zeros from MAPE/sMAPE to avoid division by zero
  nz <- actual != 0
  mape <- mean(abs((actual[nz] - predicted[nz]) / actual[nz]))
  smape <- mean(2 * abs(actual - predicted) / (abs(actual) + abs(predicted)))
  c(RMSE = rmse, MAE = mae, MAPE = mape, sMAPE = smape)
}

metrics_list <- lapply(commodities, function(comm) {
  test_ret <- commodity_data[[comm]]$test_returns
  fcast_sigma <- forecast_results[[comm]]
  len <- min(length(test_ret), length(fcast_sigma))
  m <- compute_metrics(abs(test_ret[1:len]), fcast_sigma[1:len])
  data.frame(
    Commodity = comm,
    RMSE = round(m["RMSE"], 6),
    MAE = round(m["MAE"], 6),
    MAPE = round(m["MAPE"], 6),
    sMAPE = round(m["sMAPE"], 6),
    row.names = NULL
  )
})

metrics_df <- do.call(rbind, metrics_list)
kable(metrics_df, caption = "Per-Commodity Forecast Metrics (Forecast Sigma vs |Return|)", row.names = FALSE)
```
